---
title: "STS 112: Lab 3"
author: "Your name here"
date: Due 1.29.19
output: html_notebook
---
## Getting started (1 point)
Load the `tidyverse` package and read in the data for Week 3 as a data frame called `ipums`.
```{r results = "hide"}

```
Use the `filter()` function to remove records for Alaska and Hawaii. Test that you have done this correctly.
```{r}
ipums <- ipums %>% filter()

```

## RACE and RACED (3 points)
Use the `table()` function to make a two-way (non-weighted) table of `RACED` (rows) by `RACE` (columns).
```{r}
table()
```
What is the relationship between these two variables? What does each one represent?

>

Use this week's reading to explain why IPUMS associated each variable of `RACED` with its corresponding value of `RACE`.

>

## RACED by YEAR (6 points)
Now go through all of the steps to make a high-quality weighted two-way table of `RACED` by `YEAR`. The table should look like the `RACE` by `YEAR` table we made in Notebook 2. It should have the values for `RACED` written out along the left and `YEAR` represented in columns. Numbers should have commas and no decimals, and the table should be rendered with `kable()`.
```{r}

```
Explain what this visualization *is*.

>

Explain what this visualization *shows*.

>

## Turn in your lab
Make sure you have entered your name above. Preview your lab to make sure all of your output and answers are included. Upload the `Lab3.nb.html` file to Canvas.