---
title: "STS 122 - Lab 1"
author: "Fill in your name"
output: html_notebook
---
# Lab 1: Working with Data Frames
Due 1.15.19

## Scenario
You are the proprietor of a very small bookstore. You currently have the following books:

  - *The Waste Land* by T.S. Eliot (published 1922; 10 copies in stock; $8 each)  
  - *Lord of the Flies* by William Golding (published 1954; 16 copies in stock; $8 each)  
  - *Grapes of Wrath* by John Steinbeck (published 1939; 5 copies in stock; $10 each)  
  - *Pride and Prejudice* by Jane Austen (published 1813; 7 copies in stock; $15 each)  
  - *Frankenstein* by Mary Shelley (published 1818; 15 copies in stock; $5 each)  
  - *Little Women* by Louisa May Alcott (published 1868; 9 copies in stock; $20 each)  
  - *Jane Eyre* by Charlotte Bronte (published 1847; 12 copies in stock; $25 each)  
  - *The Things They Carried* by Tim O'Brien (published 1990; 2 copies in stock; $17 each)  
  - *Huckleberry Finn* by Mark Twain (published 1884; 21 copies in stock; $12 each)  
  - *Siddhartha* by Hermann Hesse (published 1922; 4 copies in stock; $22 each)

## Inventory (1 point)
In the code block below, fill in the blanks to create a tidy data frame with a row for each book and a column for each characteristic of the book (title, author, year, etc.). We will do this by creating a vector for each characteristic, then combining the vectors into a data frame. Print your result.
```{r}
title <- c()
author <- c()
pyear <- c()
nstock <- c()
price <- c()

bookstore <- data.frame()
```

## Organization (1 point)
For the remainder of the lab, we will be using functions from `dplyr`. Load the `dplyr` package in the code block below.
```{r}

```
Saving as a new data frame called `books_sorted`, fill in the blanks below to sort your inventory by year published (most recent first, so descending order) and, within year, by author's last name (A-Z). This can be done in either one step or two, but pay attention to the order. Print your result.
```{r}
books_sorted <- bookstore %>%
```

## Sale (3 points)
Your bookstore is having a sale!  

  - Books published before 1900 are 25% off  
  - Books with more than 10 copies in stock are 40% off  
  - Books published before 1900 with more than 10 copies in stock are 50% off

In the code block below, fill in the blanks to create a new data frame with the sale price for each book, in addition to the original price. You will need three ifelse statements, which can be either sequential or nested. Print your result.
```{r}
sale <- bookstore %>% mutate()
```
In one or more complete sentences, explain what you did:

>Fill in answer here

## Sale price check (1 point)
Add a column to the data frame with the sale prices that indicates the percentage discount for each book. Do this by calculating the difference between sale price and original price. Print your result.
```{r}
sale <-
```

## Sale list (2 points)
Fill in the blanks below to reate a new data frame that includes only books on sale and only author, title, and sale price. You will need to use the `filter()` and `select()` functions, and the order you use them in matters.
```{r}
salebooks <- sale %>% %>%
```
In one or more complete sentences, explain what you did:
>

## Census and art (2 points)
In "The Census, Audiences, and Publics," Margo Anderson describes the history of two types of "census art" - nonstatistical art *about* the census and visualizations of census data. Using examples from the article and any personal experience (such as watching the clip from the West Wing in class), write a paragraph where you briefly explain what each type of art is and consider how each type of art influences the census itself.

>Fill in answer here

## Turning in this assignment
Download the `Lab1.nb.html` file from the menu on the right (check the box next to the file name), then click the `More` wheel and choose `Export`. Upload to Canvas under Lab 1.